Clinical Trial Design Using A Stopped Negative Binomial Distribution

### 2018-02-23 sii/deveaux-kane-zelterman.tex
Next version... Hopefully this goes to print.

@@

@@

The sequence of Bernoulli random variables can be modeled as the process 
$\mathbf{X} = \left\{X_k : k = 0,1,... \right\}$
with $X_0=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ \mathbbm{1}_{\{ k-t < X_k < s\}}.
\end{equation*}
where $\mathbbm{1}_{\{f\}}$ is the {\em indicator function}, taking the value 
of one if $f$ is true and zero otherwise, and
At each step a patient's outcome is measured. If it is a response, the response count increases, otherwise it stays the same. The process continues until either 
$X_k = s$ or $X_k = k-t$ corresponding to the success and failure boundaries

@@

@@

Examples of different 
shapes of the SNB are shown in in Figure~\ref{shapes.fig} varying parameters $p$, 
$s$, and $t$. The SNB distribution is a generalization of the negative 
binomial. As a result, the SNB can approximate other distributions in the same
manner as the negative binomial. When $t$ is large then $Y-s$ has a 
negative binomial distribution with
\begin{equation*}                                    %   (1)   
\mathbb{P}[Y=s+j ]        \label{nb1.eq}          
  = {{s+j-1}\choose{s-1}} p^s (1-p)^j
\end{equation*}
for $\,j=0, 1,\ldots\,$. A similar statement can be made when $s$ is large
and $t$ is moderate. As a result, with proper parameter choice, the SNB
can mimic other probability distributions in a manner similar to 
those described in \cite{Best1974} and \cite{Peizer1968}. As a generalization of the negative binomial distribution the 
SNB inherits the ability to approximate other distributions. For example,
when $s=1$ and $t \rightarrow \infty$, the SNB($p$, $s$, $t$) converges
to an negative binomial distribution with index parameter $s$ and rate parameter $p$. When $s=1$, this is the geometric distribution. The connection between the negative binomial and the gamma distribution
are well-studied in the literature (see \cite{Best1974,Ord1968,Guenther1972} for examples) as well the connection to the Poisson \cite{Anscombe1950}.

@@

@@

@@

\pagebreak

\onecolumn

\section*{Figures}

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{snb-first-plot.pdf}
\caption{
The distribution of the sample size in a trial that stops after either seven patients
respond to treatment or 11 do not when $p=0.2$.
}
\label{fig:kane_viz_1}
\end{figure}

\pagebreak

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{mean-and-variance.pdf}
\caption{
The mean and variance of the sample size for the probability of success $p$ 
when the trial stops after $s=7$ patients respond or $t=11$ fail to respond.
}
\label{fig:kane_viz_2}
\end{figure}

\pagebreak

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{KanePlot.pdf}
\caption{
A hypothetical realization of the trial.
}
\label{fig:kane_viz}
\end{figure}

\pagebreak

\begin{figure}[htbp!]
\begin{center}
\includegraphics[width=\textwidth]{likelihood.pdf}
\end{center}
\caption{Shapes of the likelihood function for given values of $Y$ with 
$s=7$ and $t=11$.
\label{likelihood.fig}}
\end{figure}

\pagebreak

\begin{figure}[htbp!]
\begin{center}
\includegraphics[width=0.8\textwidth]{shapes.pdf}
\end{center}
\caption{Different shapes of the SNB mass function with parameters 
($p$, $s$, $t$) given. Black areas indicate the mass contributed by reaching
$s$ responders before $t$ non-responders. Grey indicates
mass contributed by reaching $t$ non-responders first. \label{shapes.fig}}
\end{figure}

\pagebreak

\pagebreak

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{power-vs-significance.pdf}
\caption{
ROC curve of the all trial designs where the maximum number of patients is 17, $p=0.2$ under the null and $p=0.4$ under the alternative. The numerical values indicate the number of responses $s$ required reach the success endpoint.
}
\label{fig:all_sample_sizes_power_vs_signif}
\end{figure}

\pagebreak

\begin{figure}[htbp!]
\centering
\includegraphics[width=\textwidth]{expected-sample-size.pdf}
\caption{
The expected sample size all trial designs where the maximum number
of patients is 17 and $p=0.2$.}
\label{fig:all_sample_sizes}
\end{figure}

\pagebreak

>>>>

\begin{figure}
\centering
\includegraphics[width=0.4\textwidth]{snb-first-plot.pdf}
\caption{
The distribution of the sample size in a trial that stops after either seven patients
respond to treatment or 11 do not when $p=0.2$.
}
\label{fig:kane_viz_1}
\end{figure}

@@

\begin{figure*}
\centering
\includegraphics[width=0.7\textwidth]{mean-and-variance.pdf}
\caption{
The mean and variance of the sample size for the probability of success $p$ 
when the trial stops after $s=7$ patients respond or $t=11$ fail to respond.
}
\label{fig:kane_viz_2}
\end{figure*}

@@

The sequence of Bernoulli random variables can be modeled as the process 
$\mathbf{X} = \left\{X_k : k = 0,1,... \right\}$
with $X_0=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \ \mathbbm{1}_{\{ k-t < X_k < s\}}.
\end{equation*}
where $\mathbbm{1}_{\{f\}}$ is the {\em indicator function}, taking the value 
of one if $f$ is true and zero otherwise, and
At each step a patient's outcome is measured. If it is a response, then
the response count increases, otherwise it stays the same. The process 
continues until either 
$X_k = s$ or $X_k = k-t$ corresponding to the success and failure boundaries

\begin{figure*}%[htbp!]
\centering
\includegraphics[width=0.7\textwidth]{KanePlot.pdf}
\caption{
A hypothetical realization of the trial.
}
\label{fig:kane_viz}
\end{figure*}

@@

\begin{figure*}%[htbp!]
\centering
\includegraphics[width=0.7\textwidth]{likelihood.pdf}
\caption{Shapes of the likelihood function for given values of $Y$ with 
$s=7$ and $t=11$.}
\label{likelihood.fig}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=0.7\textwidth]{shapes.pdf}
\caption{Different shapes of the SNB mass function with parameters 
($p$, $s$, $t$) given. Black areas indicate the mass contributed by reaching
$s$ responders before $t$ non-responders. Grey indicates
mass contributed by reaching $t$ non-responders first.}
\label{fig:shapes}
\end{figure*}

Examples of different 
shapes of the SNB are shown in in Figure~\ref{fig:shapes} varying parameters $p$, 
$s$, and $t$. The SNB distribution is a generalization of the negative 
binomial. As a result, the SNB can approximate other distributions in the same
manner as the negative binomial. When $t$ is large then $Y-s$ has a 
negative binomial distribution with
\begin{equation*}                                    %   (1)   
\mathbb{P}[Y=s+j ]        \label{nb1.eq}          
  = {{s+j-1}\choose{s-1}} p^s (1-p)^j
\end{equation*}
for $\,j=0, 1,\ldots\,$. A similar statement can be made when $s$ is large
and $t$ is moderate. As a result, with proper parameter choice, the SNB
can mimic other probability distributions in a manner similar to 
those described in \cite{Best1974} and \cite{Peizer1968}. As a generalization of the negative binomial distribution the 
SNB inherits the ability to approximate other distributions. For example,
when $s=1$ and $t \rightarrow \infty$, the SNB($p$, $s$, $t$) converges
to an negative binomial distribution with index parameter $s$ and rate parameter $p$. When $s=1$, this is the geometric distribution. The connection between the negative binomial and the gamma distribution
are well-studied in the literature (see \cite{Best1974,Ord1968,Guenther1972} for examples) as well the connection to the Poisson \cite{Anscombe1950}.

@@

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.4\textwidth]{power-vs-significance.pdf}
\caption{
ROC curve of the all trial designs where the maximum number of patients is 17, $p=0.2$ under the null and $p=0.4$ under the alternative. The numerical values indicate the number of responses $s$ required reach the success endpoint.
}
\label{fig:all_sample_sizes_power_vs_signif}
\end{figure}

@@

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.4\textwidth]{expected-sample-size.pdf}
\caption{
The expected sample size all trial designs where the maximum number
of patients is 17 and $p=0.2$.}
\label{fig:all_sample_sizes}
\end{figure}

@@
